---
layout: post
title: "Mining CRAN DESCRIPTION Files"
date: 2017-05-04
output: html_document
share: true
categories: blog
excerpt: "Text analysis and more"
tags: [rstats]
---

```{r, echo = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, 
                      cache.lazy = FALSE, dpi = 180)
options(width=120, dplyr.width = 150)
library(ggplot2)
library(silgelib)
theme_set(theme_roboto())
```


A couple of weeks ago, I saw [on Dirk Eddelbuettel's blog](http://dirk.eddelbuettel.com/blog/2017/04/06/) that R 3.4.0 was going to include a function for obtaining information about packages currently on CRAN, including basically everything in DESCRIPTION files. When R 3.4.0 was released, this was one of the things I was most immediately excited about exploring, because although [I recently dabbled in scraping CRAN](http://juliasilge.com/blog/Scraping-CRAN/) to try to get this kind of information, it was rather onerous.

```{r cran}
library(tidyverse)

cran <- tools::CRAN_package_db()

# the returned data frame has two columns with the same name???
cran <- cran[,-65]

# make it a tibble
cran <- tbl_df(cran)

cran
```

There you go, all the packages currently on CRAN!

## Practices of CRAN maintainers

Some of the fields in the DESCRIPTION file of an R package tell us a bit about how a CRAN maintainer works, and in aggregate we can see how R package developers are operating.

How many packages have a URL, a place to go like GitHub to see the code and check out what is going on? 

```{r dependson="cran"}
cran %>% 
    summarise(URL = mean(!is.na(URL)))
```

What about a URL for bug reports?

```{r dependson="cran"}
cran %>% 
    summarise(BugReports = mean(!is.na(BugReports)))
```


How many packages have a package designated as a `VignetteBuilder`?

```{r dependson="cran"}
cran %>% 
    count(VignetteBuilder, sort = TRUE)
```

Are there packages that have vignettes but also have `NA` for `VignetteBuilder`? Yes, those would be packages that use [Sweave, the built-in vignette engine](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Non_002dSweave-vignettes) that comes with R. This must be biased toward older packages and it can't be a large proportion of the total, given [when CRAN has been growing the fastest](http://juliasilge.com/blog/Scraping-CRAN/). I know there are still packages with Sweave vignettes, but these days, having something in `VignetteBuilder` is at least somewhat indicative of whether a package has a vignette. There isn't anything else in the DESCRIPTION file, to my knowledge, that indicates whether a package has a vignette or not.

How many packages use testthat or RUnit for unit tests?

```{r dependson="cran"}
library(stringr)

cran %>% 
    mutate(tests = ifelse(str_detect(Suggests, "testthat|RUnit"), TRUE, FALSE),
           tests = ifelse(is.na(tests), FALSE, tests)) %>%
    summarise(tests = mean(tests))
```

(Another handful of packages have these testing suites in Imports or Depends, but not enough to change that proportion much.)

Is it the same ~20% of packages that are embracing the practices of unit tests, building vignettes, and providing a URL for bug reports?

```{r dependson="cran"}
cran %>%
    mutate(tests = ifelse(str_detect(Suggests, "testthat|RUnit"), TRUE, FALSE),
           tests = ifelse(is.na(tests), FALSE, tests),
           bug_report = ifelse(is.na(BugReports), FALSE, TRUE),
           vignette = ifelse(is.na(VignetteBuilder), FALSE, TRUE)) %>%
    count(tests, bug_report, vignette)
```

Huh, so no, actually. I would have guessed that there would have been more packages in the `TRUE/TRUE/TRUE` bin in this data frame and fewer in the bins that are mixes of `TRUE` and `FALSE`. What does that distribution look like?

```{r practices, dependson="cran", fig.width=8.5, fig.height=6}
library(tidyr)

cran %>%
    mutate(tests = ifelse(str_detect(Suggests, "testthat|RUnit"), "Tests", "No tests"),
           tests = ifelse(is.na(tests), "No tests", tests),
           bug_report = ifelse(is.na(BugReports), "No bug report", "Bug report"),
           vignette = ifelse(is.na(VignetteBuilder), "No vignette builder", "Vignette builder")) %>%
    count(tests, bug_report, vignette) %>%
    mutate(percent = n / sum(n)) %>%
    arrange(desc(percent)) %>%
    unite(practices, tests, bug_report, vignette, sep = "\n") %>%
    mutate(practices = reorder(practices, -percent)) %>%
    ggplot(aes(practices, percent, fill = practices)) +
    geom_col(alpha = 0.7, show.legend = FALSE) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(x = NULL, y = "% of CRAN pacakges",
         title = "How many packages on CRAN have units tests, a URL for bug reports, or a vignette builder?",
         subtitle = "About 6% of packages currently on CRAN have all three")
```

Maybe I should not be surprised, since a [package that I myself maintain](https://github.com/juliasilge/janeaustenr) has unit tests and a URL for bug reports but no vignette. And remember that a few of the "No vignette builder" packages are maintainers choosing to produce vignettes via Sweave, OLD SCHOOL.

## Yo dawg I heard you like Descriptions in your DESCRIPTION

One of the fields in the DESCRIPTION file for an R package is the `Description` for the package.

```{r dependson="cran"}
cran %>%
    filter(Package == "tidytext") %>%
    select(Description)
```

Let's use the [tidytext](https://github.com/juliasilge/tidytext) package that I have developed with [David Robinson](http://varianceexplained.org/) to take a look at the words maintainers use to describe their packages. What words do they use the most often?

```{r word_totals, dependson="cran"}
library(tidytext)

tidy_cran <- cran %>%
    unnest_tokens(word, Description)

word_totals <- tidy_cran %>%
    anti_join(stop_words) %>%
    count(word, sort = TRUE)

```


```{r common_words, dependson="word_totals", fig.width=7, fig.height=6}
word_totals %>%
    top_n(20) %>%
    mutate(word = reorder(word, n)) %>%
    ggplot(aes(word, n)) +
    geom_col(fill = "cyan4", alpha = 0.8) +
    coord_flip() +
    scale_y_continuous(expand = c(0,0)) +
    labs(x = NULL, y = "Number of uses in CRAN descriptions",
         title = "What are the most commonly used words in CRAN package descriptions?",
         subtitle = "After removing stop words")
```


Now let's see what the relationships between all these description words are. Let's look at how words are correlated together within description fields and make a word network.

```{r network, dependson="word_totals", fig.width=10, fig.height=7}
library(igraph)
library(ggraph)
library(widyr)

word_cors <- tidy_cran %>%
    anti_join(stop_words) %>%
    group_by(word) %>%
    filter(n() > 150) %>% # filter for words used at least 150 times
    ungroup %>%
    pairwise_cor(word, Package, sort = TRUE)

filtered_cors <- word_cors %>%
  filter(correlation > 0.2,
         item1 %in% word_totals$word,
         item2 %in% word_totals$word)

vertices <- word_totals %>%
    filter(word %in% filtered_cors$item1)

set.seed(1234)
filtered_cors %>%
    graph_from_data_frame(vertices = vertices) %>%
    ggraph(layout = "fr") +
    geom_edge_link(aes(edge_alpha = correlation), width = 2) +
    geom_node_point(aes(size = n), color = "cyan4") +
    geom_node_text(aes(label = name), repel = TRUE, point.padding = unit(0.2, "lines"),
                   family = "RobotoCondensed-Regular") +
    theme_graph(base_family = "RobotoCondensed-Regular") +
    theme(plot.title=element_text(family="Roboto-Bold")) +
    scale_size_continuous(range = c(1, 15)) +
    labs(size = "Number of uses",
         edge_alpha = "Correlation",
         title = "Word correlations in R package descriptions",
         subtitle = "Which words are more likely to occur together than with other words?")
```

## The End

If you are interested in this approach to text analysis in R, check out the [book Dave and I are publishing with O'Reilly](amazon.com/Text-Mining-R-tidy-approach/dp/1491981652/), to be released this summer, [available online as well](http://tidytextmining.com/). I found it really interesting to get a glimpse into this ecosystem that is such an important part of my professional and open-source life, both to see the overlap with the areas that I work in and the vast areas that I do not! The R Markdown file used to make this blog post is available [here](https://github.com/juliasilge/juliasilge.github.io/blob/master/_R/2017-05-04-Mining-CRAN-DESCRIPTION.Rmd). I am very happy to hear feedback and questions!

